% Scilab ( http://www.scilab.org/ ) - This file is part of Scilab
% Copyright (C) 2008-2010 - Digiteo - Michael Baudin
%
% This file must be used under the terms of the CeCILL.
% This source file is licensed as described in the file COPYING, which
% you should have received as part of this distribution.  The terms
% are also available at
% http://www.cecill.info/licences/Licence_CeCILL_V2-en.txt

\documentclass[12pt]{article}

\include{macros}

\begin{document}
\author{Micha\"el Baudin}
\date{January 2010}
\title{Introduction to Scilab}

\maketitle

\begin{abstract}
In this document, we make an overview of Scilab features so that 
we can get familiar with this environment as fast as possible.
The goal is to present the \emph{core} of 
skills necessary to start with Scilab.
In the first part, we present how to get and install this software on our 
computer. We also present how to get some help with the 
provided in-line documentation and also thanks to web resources and forums.
In the remaining sections, we present the Scilab language, especially its 
structured programming features. 
We present an important feature of Scilab, that is the management of 
real matrices and overview the linear algebra library.
The definition of functions and the elementary management of input and 
output variables is presented. 
We present Scilab's graphical features and show how to create a 2D plot,
how to configure the title and the legend and how to export that plot 
into a vectorial or bitmap format.
\end{abstract}

\tableofcontents

\input{introscilab/introscilab}

%% Bibliography

\addcontentsline{toc}{section}{Bibliography}
\bibliographystyle{plain}
\bibliography{introscilab}

\addcontentsline{toc}{section}{Index}
\printindex

\end{document}

